Typestate Inference in an Imperative Calculus

### 2012-11-30 formalism/paper.tex
replaced \sim with \precsim to make it clearer that we mean simulation rather than bisimulation (that there is a direction).

The subtyping relation $<:$ is defined in terms of a
{\it simulation preorder}. Given $O = \langle M, \Sigma, \Delta \rangle$ and
$O_2 = \langle M_2, \Sigma_2, \Delta_2 \rangle$, the simulation preorder
$\sim$
is defined to be the largest relation such that if $O@S \sim O_2@S_2$ then for all
$m$ and $S_4$ such that $(S_2, m, S_4) \in \Delta_2$ there exists an $S_3$ such that
$(S, m, S_3) \in \Delta$ and that $O@S_3 \sim O_2@S_4$.
Subtyping is defined such that $O@S <: O_2@S_2 \iff O@S \sim O_2@S_2$.

@@

It is easy to
show that $O@S \sim O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$, so these
interpretations of subtyping can be used interchangeably wherever necessary.

@@

In general, the set of
types in TS is a disjoint union of join semi-lattices.

@@

\begin{itemize}
\item $S \in O$ is the {\it state existence predicate} asserting 
$S$ is a state of $O$. 
Formally, this is equivalent to $S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
asserting the existence
of method $m$ with return type $T$ in state $S$ which triggers a 
transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the method existence predicate to state sets, meaning that for all 
$S \in \overline{S}$ there exists a state and type 
$T'$ where $(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \sim O_2@\overline{S_2}$ is the extension of the
simulation relation to support state sets, and can be interpreted as
$\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

@@

It is clear that as $\ot{O}{S} \sim \ot{O_2}{S_2}$ there must exist some
$\overline{S_3}$ such that $\ot{O}{S_3} \sim \ot{O_2}{S_4}$.
One possible solution would be to choose
$\overline{S_3} = \{ S | O@S \sim \ot{O_2}{S_4} \}$. This is not specific
enough however --- given the object protocols from \figref{fig:principalex},
$\overline{S_3} = \{ A, B, C, D \}$ as there are no possible method calls
on states $E$ or $H$ of $O_2$.

It is clear that our choice of $S_3$ must relate the set of possible method
call sequences between $\ot{O_2}{S_2}$ and $\ot{O_2}{S_4}$. For this we
define a relation between pairs of object types
$(O@\overline{S}, O_2@\overline{S_2} \lesssim 
(O@\overline{S'}, O_2@\overline{S_2'})$, which we call the
{\it simulation predecessor relation}. The definition of this relation
is given in \figref{fig:simreach}. Essentially, if $O@S \sim O_2@S_2$ and there
exists a sequence of method calls that would transition an object value of type
$O_2@\overline{S_2}$ to $O_2@\overline{S_4}$, then this sequence of method calls
would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
such that $O@\overline{S_3} \sim O@\overline{S_4}$. This is abbreviated to
$(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.

@@

\[
\begin{array}{c}
\\
\infer[\lesssim\textsc{-adj}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})}
  {
    \begin{array}{cc}
    O@\overline{S} \sim O_2@\overline{S_2}
    \\
    m : T_i \Rightarrow \overline{S_4} \in O@\overline{S_2}
    \\
    m : T_i' \Rightarrow \overline{S_3} \in O@\overline{S}
    \end{array}
  }
\\\\
\infer[\lesssim\textsc{-refl}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S}, O_2@\overline{S_2})}
  {O@\overline{S} \sim O_2@\overline{S_2}}
\\\\
\infer[\lesssim\textsc{-trans}]
  {
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  }
  {
  \begin{array}{c}
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})
  \\
  (O@\overline{S_3}, O_2@\overline{S_4}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  \end{array}
  }
\end{array}
\]
\caption{\label{fig:simreach} Definition of the simulation predecessor relation}
\end{figure}

@@

The basic types $\boolt$, $\unitt$ and $\funt{T_i}{U_i}{V}$ exhibit 
{\it structural subtyping}. However object types do not. If we treat object 
type expressions as atoms,
then a standard constraint solver which supports structural subtyping such as 
\ref{Simonet2003} can be used. This solver will find solutions for all of the 
variables that concern structural types, and leave a set of constraints between 
object type variables which we must solve through some other mechanism. 
This is the approach taken for the type constraints for TS1, which separates 
the task of {\it object constraint solving} from the main type constraint 
solving phase.

@@

\section{Soundness}

@@

\item $V = O@\overline{S'}$ for some $O$ and $\overline{S'}$. By
\remapobj, $O'@\overline{S'''} = \remapapp{T}{U}{V}$ where 
$T = O'@\overline{S''}$ and $U = O@\overline{S}$ and
$\overline{S'''} = \{ S | O'@S \sim O@\overline{S'}, S \in O' \}$ 
for some $O'$, $\overline{S}$ and $\overline{S''}$. By \subobj,
$T = O'@\overline{S'''} <: O@\overline{S'} = V$ as for each $S \in \overline{S'''}$,
$O'@S <: O@\overline{S'}$.
\end{itemize}
\end{proof}

@@

\item $V = O@\overline{S}$ for some $O$ and $\overline{S}$. 
It follows that $U = O'@\overline{S'}$ 
for some $O'$ and $\overline{S'}$ 
such that $O'@\overline{S'} \sim O@\overline{S}$ 
by rule \subobj. 
By rule \remapobj, it follows that $T = O'@\overline{S''}$ 
such that $\overline{S''} = \{ S | O'@S \sim O@\overline{S}, S \in O' \}$.
By \lemref{lem:ssetinc} it follows that 
$O'@\overline{S'} <: O'@\overline{S''}$.

@@

\begin{itemize}
\item $x = y$. It follows that $\Gamma = \Gamma_1 = \Gamma'$ and
that $T = O@\overline{S}, V = O@\overline{S'}$.
$T'$ must be of form $O'@\overline{S}''$ such that
$O'@\overline{S}'' \sim O@\overline{S}$ by rule \subobj.

Let $V' = O'@\overline{S}''' = \remapapp{T'}{T}{V}$
where
$S''' = \{ S_n' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}''\}$.
Let $U' = \bigsqcup{ \{ T_r' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}'' \} }$.
As $O'@\overline{S}'' \sim O@\overline{S}$, it follows that
$O'@\overline{S}''' \sim O@\overline{S}'$.
Therefore, $U' <: U$ and $V' <: V$. By the rule \tmethc, and \tsub
to replace $U'$ with $U$, it follows that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

>>>>

The subtyping relation $<:$ is defined in terms of a
{\it simulation preorder}. Given $O = \langle M, \Sigma, \Delta \rangle$ and
$O_2 = \langle M_2, \Sigma_2, \Delta_2 \rangle$, the simulation preorder
$\precsim$
is defined to be the largest relation such that if $O@S \precsim O_2@S_2$ then for all
$m$ and $S_4$ such that $(S_2, m, S_4) \in \Delta_2$ there exists an $S_3$ such that
$(S, m, S_3) \in \Delta$ and that $O@S_3 \precsim O_2@S_4$.
Subtyping is defined such that $O@S <: O_2@S_2 \iff O@S \precsim O_2@S_2$.

@@

It is easy to
show that $O@S \precsim O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$, so these
interpretations of subtyping can be used interchangeably wherever necessary.

@@

In general, the set of
types in TS is a join semi-lattice, with $\top$ as the greatest element.

@@

\begin{itemize}
\item $S \in O$ is the {\it state existence predicate} asserting 
$S$ is a state of $O$. 
Formally, this is equivalent to $S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
asserting the existence
of method $m$ with return type $T$ in state $S$ which triggers a 
transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the method existence predicate to state sets, meaning that for all 
$S \in \overline{S}$ there exists a state and type 
$T'$ where $(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \precsim O_2@\overline{S_2}$ is the extension of the
simulation relation to support state sets, and can be interpreted as
$\bigsqcup_{S \in \overline{S}}O@S \precsim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

@@

It is clear that as $\ot{O}{S} \precsim \ot{O_2}{S_2}$ there must exist some
$\overline{S_3}$ such that $\ot{O}{S_3} \precsim \ot{O_2}{S_4}$.
One possible solution would be to choose
$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. This is not specific
enough however --- given the object protocols from \figref{fig:principalex},
$\overline{S_3} = \{ A, B, C, D \}$ as there are no possible method calls
on states $E$ or $H$ of $O_2$.

It is clear that our choice of $S_3$ must relate the set of possible method
call sequences between $\ot{O_2}{S_2}$ and $\ot{O_2}{S_4}$. For this we
define a relation between pairs of object types
$(O@\overline{S}, O_2@\overline{S_2} \lesssim 
(O@\overline{S'}, O_2@\overline{S_2'})$, which we call the
{\it simulation predecessor relation}. The definition of this relation
is given in \figref{fig:simreach}. Essentially, if $O@S \precsim O_2@S_2$ and there
exists a sequence of method calls that would transition an object value of type
$O_2@\overline{S_2}$ to $O_2@\overline{S_4}$, then this sequence of method calls
would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
such that $O@\overline{S_3} \precsim O@\overline{S_4}$. This is abbreviated to
$(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.

@@

\[
\begin{array}{c}
\\
\infer[\lesssim\textsc{-adj}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})}
  {
    \begin{array}{cc}
    O@\overline{S} \precsim O_2@\overline{S_2}
    \\
    m : T_i \Rightarrow \overline{S_4} \in O@\overline{S_2}
    \\
    m : T_i' \Rightarrow \overline{S_3} \in O@\overline{S}
    \end{array}
  }
\\\\
\infer[\lesssim\textsc{-refl}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S}, O_2@\overline{S_2})}
  {O@\overline{S} \precsim O_2@\overline{S_2}}
\\\\
\infer[\lesssim\textsc{-trans}]
  {
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  }
  {
  \begin{array}{c}
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})
  \\
  (O@\overline{S_3}, O_2@\overline{S_4}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  \end{array}
  }
\end{array}
\]
\caption{\label{fig:simreach} Definition of the simulation predecessor relation}
\end{figure}

@@

The basic types $\boolt$, $\unitt$ and $\funt{T_i}{U_i}{V}$ exhibit 
{\it structural subtyping}. However object types do not. If we treat object 
type expressions as atoms,
then a standard constraint solver which supports structural subtyping such as 
\cite{Simonet2003} can be used. This solver will find solutions for all of the 
variables that concern structural types, and leave a set of constraints between 
object type variables which we must solve through some other mechanism. 
This is the approach taken for the type constraints for TS1, which separates 
the task of {\it object constraint solving} from the main type constraint 
solving phase.

@@

\section{Soundness}
\label{app:soundness}

@@

\item $V = O@\overline{S'}$ for some $O$ and $\overline{S'}$. By
\remapobj, $O'@\overline{S'''} = \remapapp{T}{U}{V}$ where 
$T = O'@\overline{S''}$ and $U = O@\overline{S}$ and
$\overline{S'''} = \{ S | O'@S \precsim O@\overline{S'}, S \in O' \}$ 
for some $O'$, $\overline{S}$ and $\overline{S''}$. By \subobj,
$T = O'@\overline{S'''} <: O@\overline{S'} = V$ as for each $S \in \overline{S'''}$,
$O'@S <: O@\overline{S'}$.
\end{itemize}
\end{proof}

@@

\item $V = O@\overline{S}$ for some $O$ and $\overline{S}$. 
It follows that $U = O'@\overline{S'}$ 
for some $O'$ and $\overline{S'}$ 
such that $O'@\overline{S'} \precsim O@\overline{S}$ 
by rule \subobj. 
By rule \remapobj, it follows that $T = O'@\overline{S''}$ 
such that $\overline{S''} = \{ S | O'@S \precsim O@\overline{S}, S \in O' \}$.
By \lemref{lem:ssetinc} it follows that 
$O'@\overline{S'} <: O'@\overline{S''}$.

@@

\begin{itemize}
\item $x = y$. It follows that $\Gamma = \Gamma_1 = \Gamma'$ and
that $T = O@\overline{S}, V = O@\overline{S'}$.
$T'$ must be of form $O'@\overline{S}''$ such that
$O'@\overline{S}'' \precsim O@\overline{S}$ by rule \subobj.

Let $V' = O'@\overline{S}''' = \remapapp{T'}{T}{V}$
where
$S''' = \{ S_n' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}''\}$.
Let $U' = \bigsqcup{ \{ T_r' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}'' \} }$.
As $O'@\overline{S}'' \precsim O@\overline{S}$, it follows that
$O'@\overline{S}''' \precsim O@\overline{S}'$.
Therefore, $U' <: U$ and $V' <: V$. By the rule \tmethc, and \tsub
to replace $U'$ with $U$, it follows that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
